---
title: "ghg flux plotting"
author: "Kathryn Bennett"
date: "2026-03-23"
output: html_document
---
Greenhouse gas (CO2, CH4, N2O) flux plotting
Create Figures 2, 3, S3, S4 and associated statistical analyses, and Table S5

Load packages and dataset
```{r}
library('dplyr')
library("ggplot2")
library("patchwork")
library("agricolae") #for kruskal-wallis with post-hoc test letter groupings
library("car") #for levene test, non-normal variance test
library("scales") #for plotting with log scales


#all ghg fluxes, all sampling zones
allzones_ghgflux <- read.csv(file = "data/allzones_ghgflux.csv")
summary(allzones_ghgflux)
```

format sampling_date and columns
```{r}
#format samplnig_date
allzones_ghgflux$sampling_date <- as.POSIXct(allzones_ghgflux$sampling_date, format = "%m/%d/%Y", tz = "MST")

#format sample location info as factors
allzones_ghgflux <- allzones_ghgflux %>%
  mutate(across(c(2:7), as.factor))
```

###################
Create Figure 2
###################

Setting up plotting aesthetics
```{r}
#create manual labels and ordering
zone_order <- c("tundra_undisturbed", "tundra_disturbed", "lake_disturbed","lake_center", "lake_undisturbed") #set display order of zones
labels_tz <- c("Stable tundra", "Eroding tundra", "Eroding lake shore", "Lake center", "Stable lake shore") #manual labels

#for plots of tundra only
zone_order_tundra <- c("tundra_undisturbed", "tundra_disturbed")
labels_tz_tundra <- c("Stable tundra", "Eroding tundra") #manual labels

#for plots of lake only
zone_order_lake <- c("lake_disturbed","lake_center","lake_undisturbed")
labels_tz_lake <- c("Eroding lake shore","Lake center", "Stable lake shore") #manual labels

#create customized color palette
pal_option2 <- c(tundra_undisturbed = "#FDAE6B", tundra_disturbed = "#E6550D", lake_disturbed = "#3182BD", lake_center = "#9ECAE1", lake_undisturbed = "#DEEBF7" )
```

Plot data
```{r}
#CO2 plots

# Calculate minimum y-value AND sample size for each group to add figure annotations
ann_pos <- allzones_ghgflux %>% 
  group_by(type_zone) %>% 
  summarise(
    ymin = min(co2_mg_c_m2_h, na.rm = TRUE),
    n = sum(!is.na(co2_mg_c_m2_h))   # counts non-NA co2 values
  ) %>% 
  mutate(
    x = factor(type_zone, levels = zone_order),
    y1 = ymin - 0.1 * diff(range(allzones_ghgflux$co2_mg_c_m2_h, na.rm = TRUE)),  
    y2 = ymin - 0.2 * diff(range(allzones_ghgflux$co2_mg_c_m2_h, na.rm = TRUE))  
  )
# now the plot
bp1_co2 <- ggplot(allzones_ghgflux, 
                  aes(x = factor(type_zone, levels = zone_order), 
                      y = co2_mg_c_m2_h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  geom_boxplot(aes(fill = type_zone))+
  scale_fill_manual(values = pal_option2, name = "Sampling zone",
                    breaks = zone_order, labels = labels_tz) +
  xlab(NULL) +
  ylab(expression(F[CO2]~(mg~C~m^{-2}~h^{-1}))) +

  # Kruskal-Wallis letter groupings (first annotation line)
  geom_text(data = ann_pos,
            aes(x = x, y = y1, label = c("ab","a","a","c","b")), #in alphabetical order by type_zone
            size = 4) +

  # dynamic sample counts (second annotation line)
  geom_text(data = ann_pos,
            aes(x = x, y = y2, label = n),
            fontface = "italic", size = 4) +

  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12))


## CH4 plots
#create a log scale for the plot
pseudoLog <- scales::pseudo_log_trans(base = 10)

# get data vector and transformed range
vals <- allzones_ghgflux$ch4_mg_c_m2_h
tr_min <- pseudoLog$transform(min(vals, na.rm = TRUE))
tr_max <- pseudoLog$transform(max(vals, na.rm = TRUE))
tr_range <- tr_max - tr_min

# choose an offset for annotation placement in TRANSFORMED units 
offset_tr <- 0.1 * tr_range

# compute annotation positions in data space by working in transformed space
ann_pos_ch4 <- allzones_ghgflux %>% 
  group_by(type_zone) %>% 
  summarise(
    ymin = min(ch4_mg_c_m2_h, na.rm = TRUE),
    n = sum(!is.na(ch4_mg_c_m2_h))   # counts non-NA values
  ) %>% 
  mutate(
    x = factor(type_zone, levels = zone_order),
    tr_ymin = pseudoLog$transform(ymin),
    # subtract offsets in transformed space, then invert back
    y1 = pseudoLog$inverse(tr_ymin - offset_tr),
    y2 = pseudoLog$inverse(tr_ymin - 2 * offset_tr)
  )

# Plot using those positions
bp1_ch4 <- ggplot(allzones_ghgflux, 
                  aes(x = factor(type_zone, levels = zone_order), 
                      y = ch4_mg_c_m2_h)) +
  geom_boxplot(aes(fill = type_zone)) +
  scale_fill_manual(values = pal_option2, name = "Sampling zone",
                    breaks = zone_order, labels = labels_tz) +
  xlab(NULL) +
  ylab(expression(F[CH4]~(mg~C~m^{-2}~h^{-1}))) +
  scale_y_continuous(trans = pseudoLog,
                     breaks = c(-5, -1, 0, 1, 5, 10, 20, 30)) + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +

  # Kruskal-Wallis letter groupings (first annotation line)
  geom_text(data = ann_pos_ch4,
            aes(x = x, y = y1, label = c("a","a","b","c","d")), #in alphabetical order by type_zone
            size = 4) +

  # dynamic sample sizes (second stacked annotation)
  geom_text(data = ann_pos_ch4,
            aes(x = x, y = y2, label = n),
            fontface = "italic", size = 4) +

  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12))


#CH4 data zoomed in for inset plot
bp1_ch4_zoomed <- ggplot(allzones_ghgflux, 
                  aes(x = factor(type_zone, levels = zone_order), 
                      y = ch4_mg_c_m2_h)) +
  geom_boxplot(aes(fill = type_zone))+
  scale_fill_manual(values = pal_option2, name = "Sampling zone", breaks = zone_order, labels = labels_tz) +
  xlab(NULL) +
  ylab(NULL) +
  scale_y_continuous(position = "right", limits = c(-0.1, 0.1))+
  xlim(c("tundra_undisturbed", "tundra_disturbed"))+ #axis limits to show only tundra
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "none") 

#assemble ch4 plot with zoomed tundra inset
ch4_zoomcombo <- bp1_ch4 + inset_element(bp1_ch4_zoomed, left = 0.01, bottom = 0.5, right = 0.475, top = 0.995)+
  theme(axis.text = element_text(size = 11), axis.title = element_text(size = 12),
        plot.margin = margin(t = 1, r = 2, b = 1, l = 2)  # reduce whitespace
        )


## N2O 
# Calculate minimum y-value AND sample size for each group
ann_pos_n2o <- allzones_ghgflux %>% 
  group_by(type_zone) %>% 
  summarise(
    ymin = min(n2o_mg_m2_h, na.rm = TRUE), #calc minimum
    n = sum(!is.na(n2o_mg_m2_h))   # counts non-NA values
  ) %>% 
  mutate(
    x = factor(type_zone, levels = zone_order), #create x and y coordinates for annotations
    y1 = ymin - 0.05 * diff(range(allzones_ghgflux$n2o_mg_m2_h, na.rm = TRUE)),  
    y2 = ymin - 0.08 * diff(range(allzones_ghgflux$n2o_mg_m2_h, na.rm = TRUE))  
  )

#plot
bp1_n2o <- ggplot(allzones_ghgflux, 
                  aes(x = factor(type_zone, levels = zone_order), 
                      y = n2o_mg_m2_h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_boxplot(aes(fill = type_zone))+
  scale_fill_manual(values = pal_option2, name = "Sampling zone", breaks = zone_order, labels = labels_tz) +
  xlab(NULL) +
  ylab(expression(F[N2O]~ (mg~m^{-2}~h^{-1}))) +
  scale_y_continuous(limits = c(-0.3, 0.3),
                     breaks = seq(-0.3, 0.3, by = 0.1),
                     labels = scales::number_format(accuracy = 0.1)) +
  # Kruskal-Wallis letter (first annotation line)
  geom_text(data = ann_pos_n2o,
            aes(x = x, y = y1, label = c("b","a","a","ab","ab")), #in alphabetical order by type_zone
            size = 4) +

  # dynamic sample counts (second annotation line)
  geom_text(data = ann_pos_n2o,
            aes(x = x, y = y2, label = n),
            fontface = "italic", size = 4) +
  
  scale_x_discrete(labels = labels_tz, guide = guide_axis(angle = 45)) + #rotate x-axis labels 45 deg
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1),
        axis.text = element_text(size = 11), axis.title = element_text(size = 12), 
        legend.position = "none")


#zoomed out N2O data, for inset plot
bp1_n2o_zoomout <- ggplot(allzones_ghgflux, 
                  aes(x = factor(type_zone, levels = zone_order), 
                      y = n2o_mg_m2_h)) +
  geom_boxplot(aes(fill = type_zone))+
  scale_fill_manual(values = pal_option2, name = "Sampling zone", breaks = zone_order, labels = labels_tz) +
  xlab(NULL) +
  ylab(NULL) +
  xlim(c("lake_disturbed","lake_center",  "lake_undisturbed"))+ #axis limits to show only lakes
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 11), axis.title = element_text(size = 12)) 


#assemble n2o plot with zoomed lake inset
n2o_zoomcombo <- bp1_n2o + inset_element(bp1_n2o_zoomout, left = 0.35, bottom = 0.6, right = 0.995, top = 0.995)+
  theme(axis.text = element_text(size = 11), axis.title = element_text(size = 12),
        plot.margin = margin(t = 1, r = 2, b = 1, l = 2)  # reduce whitespace
        )


#put all 3 gases together; using patchwork package
megaplot <- (bp1_co2 / ch4_zoomcombo / n2o_zoomcombo )+
  plot_layout(guides = "auto") 
megaplot #print plot

#export plot
ggsave(megaplot, 
       filename = "plots/ghgflux_byzone.png",
       device = "png",
       height = 9, width = 6.5, units = "in") 
```

Statistical analyses of box plot groupings (Figure 2)
```{r}
## CO2
#shapiro test to assess normality
with(allzones_ghgflux, tapply( co2_mg_c_m2_h, type_zone, shapiro.test)) #test each type_zone group
# Levene's test for homogeneity of variance on non-normal data
leveneTest(co2_mg_c_m2_h ~ type_zone, data = allzones_ghgflux)
#Kruskal-wallis test for non-parametric data
kruskal(allzones_ghgflux$co2_mg_c_m2_h, allzones_ghgflux$type_zone, p.adj =c("bonferroni"), group = TRUE, console = TRUE)

## CH4
#shapiro test to assess normality
with(allzones_ghgflux, tapply( ch4_mg_c_m2_h, type_zone, shapiro.test)) #test each type_zone group
# Levene's test for homogeneity fo variance on non-normal data
leveneTest(ch4_mg_c_m2_h ~ type_zone, data = allzones_ghgflux)
#Kruskal-wallis test for non-parametric data
kruskal(allzones_ghgflux$ch4_mg_c_m2_h, allzones_ghgflux$type_zone, p.adj =c("bonferroni"), group = TRUE, console = TRUE)

## N2O
#shapiro test to assess normality
with(allzones_ghgflux, tapply( n2o_mg_m2_h, type_zone, shapiro.test))
# Levene's test for homogeneity of variance on non-normal data
leveneTest(n2o_mg_m2_h ~ type_zone, data = allzones_ghgflux)
#Kruskal-wallis for non-parametric data
kruskal(allzones_ghgflux$n2o_mg_m2_h, allzones_ghgflux$type_zone, p.adj =c("bonferroni"), group = TRUE, console = TRUE)
```

#############################################
Figure S4 - Lake FN2O with hot moment removed
##############################################

Statistical analysis
```{r}
#remove N2O hot moment on 31 July 2023
n2o_noLakeHM <- allzones_ghgflux %>% 
  filter(sampling_date != as.POSIXct("2023-07-31", tz = "MST", format = "%Y-%m-%d")) %>% 
           filter(flux_type == "lake") #lake data only
#check output - note not in date order
unique(n2o_noLakeHM$sampling_date)

#shapiro test for normality
with(n2o_noLakeHM, tapply( n2o_mg_m2_h, type_zone, shapiro.test))
# Levene's test for homogeneity fo variance on non-normal data
leveneTest(n2o_mg_m2_h ~ type_zone, data = n2o_noLakeHM)
#Kruskal-Wallis test
kruskal(n2o_noLakeHM$n2o_mg_m2_h, n2o_noLakeHM$type_zone, p.adj =c("bonferroni"), group = TRUE, console = TRUE)
```

Create plot - Figure S4
```{r}
#plot
bp_n2ohm <- ggplot(n2o_noLakeHM, 
                  aes(x = factor(type_zone, levels = zone_order_lake), 
                      y = n2o_mg_m2_h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_boxplot(aes(fill = type_zone))+
  scale_fill_manual(values = pal_option2, name = "Sampling zone", breaks = zone_order_lake, labels = labels_tz_lake) +
  xlab(NULL) +
  xlim(c("lake_disturbed","lake_center",  "lake_undisturbed"))+ #axis limits to show only lake
  ylab(expression(F[N2O]~ (mg~m^{-2}~h^{-1}))) +
  scale_y_continuous(limits = c(-0.3, 0.3),
                     breaks = seq(-0.3, 0.3, by = 0.1),
                     labels = scales::number_format(accuracy = 0.1)) +
  scale_x_discrete(labels = labels_tz_lake, guide = guide_axis(angle = 45)) + #rotate x-axis labels 45 deg
  annotate("text", x = c(1,2,3), y = -0.2, label = c("a", "b", "ab"), size = 4)+
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1),
        axis.text = element_text(size = 11), axis.title = element_text(size = 12), 
        legend.position = "none")
bp_n2ohm #print plot

#export plot
ggsave(bp_n2ohm,
       filename = "plots/laken2o_noHM.png",
       device = "png",
       height = 5, width = 5, units = "in")
```

############################################### 
Figure S3 - Tundra CO2 component fluxes (ER, GPP)
###############################################

Statistical analyses of differences between sampling zones
```{r}
#Filter to tundra data only
tundra_ghgflux <- allzones_ghgflux %>%
  filter(flux_type == "tundra")

## ER
#shapiro test for normality
with(tundra_ghgflux, tapply( er_mg_c_m2_h, type_zone, shapiro.test))
# Levene's test for homogeneity fo variance on non-normal data
leveneTest(er_mg_c_m2_h ~ type_zone, data = tundra_ghgflux)
#mann-whitney u test for non-parametric data
wilcox.test(er_mg_c_m2_h ~ type_zone, data = tundra_ghgflux)

## GPP
#shapiro test for normality
with(tundra_ghgflux, tapply( gpp_mg_c_m2_h, type_zone, shapiro.test))
# Levene's test for homogeneity fo variance on non-normal data
leveneTest(gpp_mg_c_m2_h ~ type_zone, data = tundra_ghgflux)
#mann-whitney u test for non-parametric data
# H1: tundra_disturbed (group 1) is more negative - aka less than undisturbed
wilcox.test(gpp_mg_c_m2_h ~ type_zone, data = tundra_ghgflux, alternative = "less")
```

Create Figure S3
```{r}
## ER
# Calculate minimum y-value to place annotation AND sample size for each group
ann_pos_er <- tundra_ghgflux %>% 
  group_by(type_zone) %>% 
  summarise(
    ymin = min(er_mg_c_m2_h, na.rm = TRUE),
    n = sum(!is.na(er_mg_c_m2_h))   # counts non-NA co2 values
  ) %>% 
  mutate(
    x = factor(type_zone, levels = zone_order_tundra),
    y1 = ymin - 0.1 * diff(range(tundra_ghgflux$er_mg_c_m2_h, na.rm = TRUE))  
  )

#plot
tundra_er <- ggplot(tundra_ghgflux, 
                  aes(x = factor(type_zone, levels = zone_order_tundra), 
                      y = er_mg_c_m2_h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  geom_boxplot(aes(fill = type_zone))+
  scale_fill_manual(values = pal_option2, name = "Sampling zone", breaks = zone_order_tundra, labels = labels_tz_tundra) +
  xlab(NULL) +
  ylab(expression(ER~(mg~C~m^{-2}~h^{-1}))) + 
  scale_y_continuous(breaks = seq(-500, 300, by = 100), limits = c(-600, 300))+
  geom_text(data = ann_pos_er, #annotation line (dynamic sample counts)
            aes(x = x, y = y1, label = n),
            fontface = "italic", size = 4) +
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "bottom",
        axis.text = element_text(size = 11), axis.title = element_text(size = 12)) 

## GPP
# Calculate minimum y-value to place annotation AND sample size for each group
ann_pos_gpp <- tundra_ghgflux %>% 
  group_by(type_zone) %>% 
  summarise(
    ymin = min(gpp_mg_c_m2_h, na.rm = TRUE),
    n = sum(!is.na(gpp_mg_c_m2_h))   # counts non-NA co2 values
  ) %>% 
  mutate(
    x = factor(type_zone, levels = zone_order_tundra),
    y1 = ymin - 0.1 * diff(range(tundra_ghgflux$gpp_mg_c_m2_h, na.rm = TRUE))  
  )

#plot
tundra_gpp <- ggplot(tundra_ghgflux, 
                  aes(x = factor(type_zone, levels = zone_order_tundra), 
                      y = gpp_mg_c_m2_h)) +
  geom_boxplot(aes(fill = type_zone))+
  scale_fill_manual(values = pal_option2, name = "Sampling zone", breaks = zone_order_tundra, labels = labels_tz_tundra) +
  xlab(NULL) +
  ylab(expression(GPP~(mg~C~m^{-2}~h^{-1}))) + 
  scale_y_continuous(breaks = seq(-500, 300, by = 100), limits = c(-600, 300))+
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  annotate("text", x = 1.5, y = 225, label = c("*"), size = 12)+ #mark significant difference
  geom_text(data = ann_pos_gpp, #annotation line (dynamic sample counts)
            aes(x = x, y = y1, label = n),
            fontface = "italic", size = 4) +
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 11), axis.title = element_text(size = 12)) 

#put panels together; using patchwork package
components_co2plot <- (tundra_er + tundra_gpp )+
  plot_layout(guides = "collect")& 
  theme(legend.position = 'bottom', legend.text = element_text(size = 11),legend.title = element_text(size = 12))
components_co2plot #print plot

#export plot
ggsave(components_co2plot, 
       filename = "plots/componentsco2_boxplot.png",
       device = "png",
       height = 4.5, width = 6.5, units = "in")
```

###############################
Table S5 - Summary of ghg fluxes
###############################

```{r}
#Summary flux data tables - calculate mean, sd, median, quartiles 
#group by sampling zone
sumtab_byzone <- allzones_ghgflux %>% 
  dplyr::group_by(type_zone) %>% 
  summarise(across(
    .cols = where(is.numeric), 
    .fns = list(mean = ~mean(.x, na.rm = T), 
                sd = ~sd(.x, na.rm = T), 
                med = ~median(.x, na.rm = T),
                q1 = ~quantile(.x, probs = 0.25, na.rm = T),
                q3 = ~quantile(.x, probs = 0.75, na.rm = T)),
    .names = "{col}_{fn}"
  )) 
sumtab_byzone

#group by lakes v tundra (all sampling zones)
sumtab_bysystem <- allzones_ghgflux %>% 
  dplyr::group_by(flux_type) %>% 
  summarise(across(
    .cols = where(is.numeric), 
    .fns = list(mean = ~mean(.x, na.rm = T), 
                sd = ~sd(.x, na.rm = T), 
                med = ~median(.x, na.rm = T),
                q1 = ~quantile(.x, probs = 0.25, na.rm = T),
                q3 = ~quantile(.x, probs = 0.75, na.rm = T)),
    .names = "{col}_{fn}"
  )) 
sumtab_bysystem
```

########################################
Figure 3 - Vegetated vs. bare soil fluxes
#######################################

```{r}
#use data subset of tundra zones only
#add new var to easily plot barren, veg, and stable
tundra_ghgflux$zone_veg <- paste(tundra_ghgflux$sampling_zone, tundra_ghgflux$vegetated, sep = "_")

#setup plot aestherics
#manual labeling order for tundra
zone_order_barren <- c("undisturbed_y", "disturbed_y", "disturbed_n")
labels_barren <- c("vegetated stable", "vegetated eroding", "bare soil eroding") #manual labels

#color palette
pal_veg <- c(undisturbed_y = "lightgreen", disturbed_y = "darkgreen", disturbed_n = "brown" )

# 3 panel plot to show FCO2 vs ER vs GPP

## FCO2
veg_fco2 <- ggplot(tundra_ghgflux, 
                  aes(x = factor(zone_veg, levels = zone_order_barren), 
                      y = co2_mg_c_m2_h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  geom_boxplot(aes(fill = zone_veg))+
  scale_fill_manual(values = pal_veg, name = "Sampling zone", breaks = zone_order_barren, labels = labels_barren) +
  xlab(NULL) +
  ylab(expression(F[CO2]~(mg~C~m^{-2}~h^{-1}))) + 
  scale_y_continuous(breaks = seq(-500, 300, by = 100), limits = c(-500, 300))+
  annotate("text", x = c(1,2,3), y = 275, label = c("a", "b","a"), size = 4)+ #Kruskal Wallis groupings
  annotate("text", x = c(1,2,3), y = -500, label = c("101", "53","36"), fontface = "italic", size = 4)+ #sample count, manually determined
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 11), axis.title = element_text(size = 12)) 

## ER
veg_er <- ggplot(tundra_ghgflux, 
                  aes(x = factor(zone_veg, levels = zone_order_barren), 
                      y = er_mg_c_m2_h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  geom_boxplot(aes(fill = zone_veg))+
  scale_fill_manual(values = pal_veg, name = "Tundra type", breaks = zone_order_barren, labels = labels_barren) +
  xlab(NULL) +
  ylab(expression(ER~(mg~C~m^{-2}~h^{-1}))) + 
  scale_y_continuous(breaks = seq(-500, 300, by = 100), limits = c(-500, 300))+
  annotate("text", x = c(1,2,3), y = 275, label = c("b", "a","c"), size = 4)+ #Kruskal Wallis groupings
  annotate("text", x = c(1,2,3), y = -500, label = c("100", "62","37"),fontface = "italic", size = 4)+ #manual sample counts
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "none", 
        axis.text = element_text(size = 11), axis.title = element_text(size = 12),
        legend.title = element_text(size = 12), legend.text = element_text(size = 11)) 

## GPP
veg_gpp <- ggplot(tundra_ghgflux, 
                  aes(x = factor(zone_veg, levels = zone_order_barren), 
                      y = gpp_mg_c_m2_h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  geom_boxplot(aes(fill = zone_veg))+
  scale_fill_manual(values = pal_veg, name = "Sampling zone", breaks = zone_order_barren, labels = labels_barren) +
  xlab(NULL) +
  ylab(expression(GPP~(mg~C~m^{-2}~h^{-1}))) + 
  scale_y_continuous(breaks = seq(-500, 300, by = 100), limits = c(-500, 300))+
  annotate("text", x = c(1,2,3), y = 275, label = c("b", "c","a"), size = 4)+ #Kruskal Wallis groupings
  annotate("text", x = c(1,2,3), y = -500, label = c("92", "50","34"), fontface = "italic", size = 4)+ #manual sample counts
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 11), axis.title = element_text(size = 12)) 

## CH4
veg_ch4 <- ggplot(tundra_ghgflux, 
                  aes(x = factor(zone_veg, levels = zone_order_barren), 
                      y = ch4_mg_c_m2_h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  geom_boxplot(aes(fill = zone_veg))+
  scale_fill_manual(values = pal_veg, name = "Sampling zone", breaks = zone_order_barren, labels = labels_barren) +
  xlab(NULL) +
  ylab(expression(F[CH4]~(mg~C~m^{-2}~h^{-1}))) + 
  annotate("text", x = c(1,2,3), y = 0.1, label = c("b","a","a"), size = 4)+ #Kruskal Wallis groupings
  annotate("text", x = c(1,2,3), y = -0.15, label = c("111", "67","39"),fontface = "italic", size = 4)+ #manual sample count
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 11), axis.title = element_text(size = 12)) 

## N2O
veg_n2o <- ggplot(tundra_ghgflux, 
                  aes(x = factor(zone_veg, levels = zone_order_barren), 
                      y = n2o_mg_m2_h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  geom_boxplot(aes(fill = zone_veg))+
  scale_fill_manual(values = pal_veg, name = "Microsite & \nsampling zone", breaks = zone_order_barren, labels = labels_barren) +
  xlab(NULL) +
  ylab(expression(F[N2O]~(mg~m^{-2}~h^{-1}))) +
  annotate("text", x = c(1,2,3), y = -0.15, label = c("8", "4","3"),fontface = "italic", size = 4)+ #manual sample count
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 11), axis.title = element_text(size = 12),
        legend.title = element_text(size = 12), legend.text = element_text(size = 11)) 

## combine all 5 plots; using cowplot package
library(cowplot)
# Extract a shared legend from one of the plots
shared_legend <- get_legend(
  veg_n2o + theme(legend.position = "right")  # pick one plot for legend
)
# Turn the legend into a plot
veg_legend_plot <- ggdraw(shared_legend)

veg_all_plots <- plot_grid(
  veg_fco2, veg_er, veg_gpp,
  veg_ch4, veg_n2o, veg_legend_plot,
  ncol = 3,
  align = "hv",
  rel_heights = c(1, 1),
  rel_widths = c(1, 1, 1)
)
veg_all_plots

#export plot
#note: some text alignment issues in the legend, fixed in an image editing software
ggsave(veg_all_plots, 
       filename = "plots/veg_allfx_boxplot.png",
       device = "png",
       height = 5, width = 6.5, units = "in")
```

statistical analyses of fluxes by vegetated (vegetated vs. bare soil) groupings
```{r}
#format zone_veg as a factor
tundra_ghgflux$zone_veg <- as.factor(tundra_ghgflux$zone_veg)

## FCO2 
#shapiro test for normality
with(tundra_ghgflux, tapply( co2_mg_c_m2_h, zone_veg, shapiro.test))
# Levene's test for homogeneity of variance on non-normal data
leveneTest(co2_mg_c_m2_h ~ zone_veg, data = tundra_ghgflux)
#kruskal-wallis test
kruskal(tundra_ghgflux$co2_mg_c_m2_h, tundra_ghgflux$zone_veg, p.adj =c("bonferroni"), group = TRUE, console = TRUE)

## ER
#shapiro test for normality
with(tundra_ghgflux, tapply( er_mg_c_m2_h, zone_veg, shapiro.test))
# Levene's test for homogeneity fo variance on non-normal data
leveneTest(er_mg_c_m2_h ~ zone_veg, data = tundra_ghgflux)
#kruskal-wallis
kruskal(tundra_ghgflux$er_mg_c_m2_h, tundra_ghgflux$zone_veg, p.adj =c("bonferroni"), group = TRUE, console = TRUE)

## GPP
#shapiro test for normality
with(tundra_ghgflux, tapply( gpp_mg_c_m2_h, zone_veg, shapiro.test))
# Levene's test for homogeneity fo variance on non-normal data
leveneTest(gpp_mg_c_m2_h ~ zone_veg, data = tundra_ghgflux)
#kruskal-wallis
kruskal(tundra_ghgflux$gpp_mg_c_m2_h, tundra_ghgflux$zone_veg, p.adj =c("bonferroni"), group = TRUE, console = TRUE)

## CH4
#shapiro test for normality
with(tundra_ghgflux, tapply( ch4_mg_c_m2_h, zone_veg, shapiro.test))
# Levene's test for homogeneity fo variance on non-normal data
leveneTest(ch4_mg_c_m2_h ~ zone_veg, data = tundra_ghgflux)
#kruskal-wallis
kruskal(tundra_ghgflux$ch4_mg_c_m2_h, tundra_ghgflux$zone_veg, p.adj =c("bonferroni"), group = TRUE, console = TRUE)

## not testing N2O, sample size in groups too small (n = 3, 4, 8)
```
