---
title: "Tundra and lake LMMs"
author: "Kathryn Bennett"
date: "2026-03-24"
output: html_document
---

Linear mixed-effects models (LMMs) on tundra CO2 and CH4 fluxes
& lake CO2, CH4, N2O fluxes

load packages and data
```{r}
library("dplyr")
library("lme4")
library("sjPlot")

tundra_lmmdata <- read.csv(file = "data/tundra_lmmdata.csv")
lake_lmmdata <- read.csv(file = "data/lake_lmmdata.csv")
```

#################
Tundra LMMs
#################

Visually assess normality
```{r}
hist(tundra_lmmdata$nee_mg_c_m2_h)
qqnorm(tundra_lmmdata$nee_mg_c_m2_h)
qqline(tundra_lmmdata$nee_mg_c_m2_h, col = "red")

hist(tundra_lmmdata$ch4_mg_c_m2_h)
qqnorm(tundra_lmmdata$ch4_mg_c_m2_h)
qqline(tundra_lmmdata$ch4_mg_c_m2_h, col = "red")
```


FCO2 LMM
```{r}
#LMM
og_lmm_nee <- lmer(nee_mg_c_m2_h ~  sampling_zone + vegetated +
    vwc_perc_mean + frost_cm_mean + average_par_out + green_perc + average_t_out + (1 | site/collar_id), data = tundra_lmmdata)
summary(og_lmm_nee)

#calculate % variance explained by random effects - see random effects table in summary()
#new vwc data
1833/(1833+463+3008)*100 #34% explained by collar_id:site
463/(1833+463+3008)*100 #9% explained by site alone
3008/(1833+463+3008)*100 #57% explained by fixed effects

#examine residuals
par(mfrow = c(1, 1))
plot(og_lmm_nee, col = "black", xlab = expression("fitted tundra F"[CO2]~"values (mg C m"^-2~"h"^-1*")"), ylab = "residuals")
qqnorm(resid(og_lmm_nee), main = "Q-Q plot")
qqline(resid(og_lmm_nee), col = "red") #good, fit straight line

#visualize table of results
tab_model(og_lmm_nee,
          show.p = FALSE, #no p-values, calculated below with Likelihood Ratios
  show.stat = TRUE, #t value  
  show.df = TRUE #show degrees of freedom
  )
```

FCO2 continued: calcualte p-values using likelihood ratio tests (ANOVA)
```{r}
## Likelihood ratio tests: testing removal of each variable to determine significance 

#full model, note: REML = F
Form1 <- formula(nee_mg_c_m2_h ~  sampling_zone + vegetated + vwc_perc_mean + frost_cm_mean + average_par_out + green_perc + average_t_out + (1 | site/collar_id))
M1.Full <- lmer(Form1, data = tundra_lmmdata, REML = F)

#test removing each fixed effect 

M1.A <- update(M1.Full, .~. -sampling_zone) #remove sampling_zone 
anova(M1.Full, M1.A) #compare to full model
  
M1.B <- update(M1.Full, .~. -vegetated) #remove vegetated
anova(M1.Full, M1.B) #compare to full model

#remove vwc NAs to compare datasets of the same size
tundra_novwcNA <- tundra_lmmdata[!is.na(tundra_lmmdata$vwc_perc_mean), ]
M1.Full.vwc <- lmer(nee_mg_c_m2_h ~  sampling_zone + vegetated + vwc_perc_mean + frost_cm_mean + average_par_out + green_perc + average_t_out + (1 | site/collar_id),
             data = tundra_novwcNA, REML = F)
M1.C <- update(M1.Full.vwc, .~. -vwc_perc_mean) #remove vwc
anova(M1.Full.vwc, M1.C) #compare to full model

#frost impacts data size, remove all NA rows from frost to compare 2 dfs of the same size
tundra_nofrostNA <- tundra_lmmdata[!is.na(tundra_lmmdata$frost_cm_mean), ]
M1.Full.frost <- lmer(nee_mg_c_m2_h ~  sampling_zone + vegetated + vwc_perc_mean + frost_cm_mean + average_par_out + green_perc + average_t_out + (1 | site/collar_id),
             data = tundra_nofrostNA, REML = F)
M1.D <- update(M1.Full.frost, .~. -frost_cm_mean)
anova(M1.Full.frost, M1.D) #compare to full model

M1.E <- update(M1.Full, .~. -average_par_out) #remove par
anova(M1.Full, M1.E)
  #not significant p = 0.4936, X2 = 0.4686
M1.F <- update(M1.Full, .~. -average_t_out) #remove air T
anova(M1.Full, M1.F) #compare to full model

# removing NAs for greenness to compare 2 dfs of the same size
tundra_nogreenNA <- tundra_lmmdata[!is.na(tundra_lmmdata$green_perc), ]
M1.Full.green <- lmer(nee_mg_c_m2_h ~  sampling_zone + vegetated + vwc_perc_mean + frost_cm_mean + average_par_out + green_perc + average_t_out + (1 | site/collar_id),
             data = tundra_nogreenNA, REML = F)

M1.G <- update(M1.Full.green, .~. -green_perc) #remove greenness
anova(M1.Full.green, M1.G) #compare to full model

```


CH4 LMM
```{r}
#LMM
lmm_ch4 <- lmer(ch4_mg_c_m2_h ~  sampling_zone + vegetated +
    vwc_perc_mean + frost_cm_mean + average_t_out + green_perc + (1 | site/collar_id), data = tundra_lmmdata)

summary(lmm_ch4)

#examine residuals
plot(lmm_ch4, col = "black", xlab = expression("fitted tundra F"[CH4]~"values (mg C m"^-2~"h"^-1*")"), ylab = "residuals")
qqnorm(resid(lmm_ch4), main = "Q-Q plot")
qqline(resid(lmm_ch4), col = "red") 

#relative % of variance explained by random effects
(2.296e-05/(2.296e-05+3.345e-05+3.459e-04))*100 #6% explained by collar_id:site
(3.345e-05/(2.296e-05+3.345e-05+3.459e-04))*100 #8% explained by site
(3.459e-04/(2.296e-05+3.345e-05+3.459e-04))*100 #86% explained by fixed effects

#visualize table of results
tab_model(lmm_ch4,
          show.p = FALSE, 
  show.stat = TRUE, #t value  
  show.df = TRUE, #show degrees of freedom
 digits = 4) #digits after the decimal
```

Calculate p-values
```{r}
#likelihood ratio tests

#full model
tch4.form1 <- formula(ch4_mg_c_m2_h ~  sampling_zone + vegetated +
    vwc_perc_mean + frost_cm_mean + green_perc + average_t_out + (1 | site/collar_id))
tch4.full <- lmer(tch4.form1, data = tundra_lmmdata, REML = F)

#test removing each fixed effect 

tch4.a <- update(tch4.full, .~. -sampling_zone) #remove sampling zone
anova(tch4.full, tch4.a) #compare to full model

tch4.b <- update(tch4.full, .~. -vegetated) #vegetated y or n
anova(tch4.full, tch4.b) #compare to full model
 
# removing NAs of VWC to compare dfs of the same size
tundra_novwcNA <- tundra_lmmdata[!is.na(tundra_lmmdata$vwc_perc_mean), ]
tch4.full.vwc <- lmer(ch4_mg_c_m2_h ~  sampling_zone + vegetated +
    vwc_perc_mean + frost_cm_mean + green_perc + average_t_out + (1 | site/collar_id), data = tundra_novwcNA, REML = F)
tch4.c <- update(tch4.full.vwc, .~. -vwc_perc_mean) #remove VWC
anova(tch4.full.vwc, tch4.c) #compare to full model

# removing NAs of frost table to compare dfs of the same size
tundra_nofrostNA <- tundra_lmmdata[!is.na(tundra_lmmdata$frost_cm_mean), ]
tch4.full.frost <- lmer(ch4_mg_c_m2_h ~  sampling_zone + vegetated +
    vwc_perc_mean + frost_cm_mean + green_perc + average_t_out + (1 | site/collar_id), data = tundra_nofrostNA, REML = F)
tch4.d <- update(tch4.full.frost, .~. -frost_cm_mean) #remove frost table
anova(tch4.full.frost, tch4.d) #compare to full model

#remove NAs from greenness to compare dfs of the same size
tundra_nogreenNA <- tundra_lmmdata[!is.na(tundra_lmmdata$green_perc), ]
tch4.full.green <- lmer(ch4_mg_c_m2_h ~  sampling_zone + vegetated + vwc_perc_mean + frost_cm_mean + green_perc + average_t_out + (1 | site/collar_id),
             data = tundra_nogreenNA, REML = F)
tch4.e <- update(tch4.full.green, .~. -green_perc) #remove greenness
anova(tch4.full.green, tch4.e) #compare to full model


tch4.f <- update(tch4.full, .~. -average_t_out) #remove air temp
anova(tch4.full, tch4.f) #compare to full model

```

#################
Lake LMMs
#################

Lake FCO2 LMM
```{r}
#LMM for lake CO2
co2_lake_m1 <- lmer(co2_mg_c_m2_h ~ tsurf_water_degc + cond_uscm.1 + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no), data = lake_lmmdata)
summary(co2_lake_m1) 

#examine residuals
plot(co2_lake_m1, col = "black", xlab = expression("fitted lake F"[CO2]~"values (mg C m"^-2~"h"^-1*")"), ylab = "residuals")
qqnorm(resid(co2_lake_m1), main = "Q-Q plot")
qqline(resid(co2_lake_m1), col = "red") #good, fit straight line

#% variance calculation
  #0 % variance explained by individual plots
(40.57/(40.57+141.1))*100 #22% explained by zone (center, undist, dist); 
  #0% explained by site aka lake
(141.1/(40.57+141.1))*100 #78% explained by fixed effects

#visualize table of outputs
tab_model(co2_lake_m1,
          show.p = FALSE, #no p-values, calculated with likelihood ratio 
  show.stat = TRUE, #t value  
  show.df = TRUE, #show degrees of freedom
 digits = 2)

```

likelihood ratio tests - co2 model
note: see tundra LRTs for more thorough commenting explaining code steps
```{r}
#full model, note: REML = F
Form1 <- formula(co2_mg_c_m2_h ~ tsurf_water_degc + cond_uscm.1 + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no))
Lco2.full <- lmer(Form1, data = lake_lmmdata, REML = F)

#remove each fixed effect

Lco2.A <- update(Lco2.full, .~. -tsurf_water_degc) #remove twater
anova(Lco2.full, Lco2.A)

#remove EC NAs to compare dfs of the same size
lake_noEC <- lake_lmmdata[!is.na(lake_lmmdata$cond_uscm.1),]
Lco2.full.ec <- lmer(co2_mg_c_m2_h ~ tsurf_water_degc + cond_uscm.1 + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no), data = lake_noEC, REML = F)
Lco2.B <- update(Lco2.full.ec, .~. -cond_uscm.1) #remove EC
anova(Lco2.full.ec, Lco2.B)
  
#remove DO NAs to compare dfs of the same size
lake_nodoNA <- lake_lmmdata[!is.na(lake_lmmdata$do_mgl.1),]
Lco2.full.do <- lmer(co2_mg_c_m2_h ~ tsurf_water_degc + cond_uscm.1 + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no), data = lake_nodoNA, REML = F)
Lco2.D <- update(Lco2.full.do, .~. -do_mgl.1) #remove DO
anova(Lco2.full.do, Lco2.D)
  

Lco2.E <- update(Lco2.full, .~. -disturbance) #remove disturbance
anova(Lco2.full, Lco2.E)
  
Lco2.F <- update(Lco2.full, .~. -area_km2) #remove area
anova(Lco2.full, Lco2.F)

Lco2.G <- update(Lco2.full, .~. -depth_m) #remove depth
anova(Lco2.full, Lco2.G)

```


LMM on lake CH4

```{r}
hist(lake_lmmdata$ch4_mg_c_m2_h)
#transform ch4
lake_lmmdata$ch4_cuberoot <- sign(lake_lmmdata$ch4_mg_c_m2_h) * abs(lake_lmmdata$ch4_mg_c_m2_h)^(1/3)
print(lake_lmmdata$ch4_cuberoot)
hist(lake_lmmdata$ch4_cuberoot)
```

CH4, Model 1
```{r}
#model 1
ch4_lake_m1 <- lmer(ch4_cuberoot ~ tsurf_water_degc + cond_uscm.1 + ph + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no), data = lake_lmmdata)
summary(ch4_lake_m1) 

#examine residuals
plot(ch4_lake_m1, col = "black",xlab = expression(atop("fitted lake (F"[CH4]*")"^(1/3)~"values",
  "(mg C m"^-2~"h"^-1*") from model 1")), 
  ylab = "residuals" )
qqnorm(resid(ch4_lake_m1), main = "Q-Q plot")
qqline(resid(ch4_lake_m1), col = "red")

#variance percentages
((0.039242)/(0.039242+0.286660+0.006439+0.335162))*100 #6% explained by chamber no
((0.286660)/(0.039242+0.286660+0.006439+0.335162))*100 # 43% explained by sampling zone
((0.006439)/(0.039242+0.286660+0.006439+0.335162))*100 # 1% explained by lake
((0.335162)/(0.039242+0.286660+0.006439+0.335162))*100 #50% explained by fixed effects


#visualize table of outputs **data still transformed**
tab_model(ch4_lake_m1,
          show.p = FALSE, #no p-values 
  show.stat = TRUE, #t value  
  show.df = TRUE, #show degrees of freedom
 digits = 4)
```

likelihood ratios for CH4 Model 1 aka ch4_lake_m1
```{r}
#full model
Form2 <- formula(ch4_cuberoot ~ tsurf_water_degc + cond_uscm.1 + ph + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no))
M1ch4.full <- lmer(Form2, data = lake_lmmdata, REML = F)

# test removal of fixed effects

lake_nophNA <- lake_lmmdata[!is.na(lake_lmmdata$ph),]
M1ch4.full.ph <- lmer(ch4_cuberoot ~ tsurf_water_degc + cond_uscm.1 + ph + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no), data = lake_nophNA, REML = F)
M1ch4.A <- update(M1ch4.full.ph, .~. -ph) #remove ph
anova(M1ch4.full.ph, M1ch4.A)

lake_nodoNA <- lake_lmmdata[!is.na(lake_lmmdata$do_mgl.1),]
M1ch4.full.do <- lmer(ch4_cuberoot ~ tsurf_water_degc + cond_uscm.1 + ph + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no), data = lake_nodoNA, REML = F)
M1ch4.B <- update(M1ch4.full.do, .~. -do_mgl.1) #remove do
anova(M1ch4.full.do, M1ch4.B)

M1ch4.C <- update(M1ch4.full, .~. -tsurf_water_degc) #remove Twater
anova(M1ch4.full, M1ch4.C)

M1ch4.D <- update(M1ch4.full, .~. -cond_uscm.1) #remove conductivity
anova(M1ch4.full, M1ch4.D)

M1ch4.E <- update(M1ch4.full, .~. -disturbance) #remove disturbance
  anova(M1ch4.full, M1ch4.E)

M1ch4.F <- update(M1ch4.full, .~. -area_km2) #remove area
  anova(M1ch4.full, M1ch4.F)

M1ch4.G <- update(M1ch4.full, .~. -depth_m) #remove depth
  anova(M1ch4.full, M1ch4.G)

```


Lake CH4, Model 2 LMM

```{r}
#second viable model, removes DO (NA = 72), ph (NA = 52) to use larger dataset. Keeping EC, different sensor on pH probe
#119 observations in previous model, 206 in this model

ch4_lake_m2 <- lmer(ch4_cuberoot ~ tsurf_water_degc + cond_uscm.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no), data = lake_lmmdata)
summary(ch4_lake_m2)

#examine residuals
plot(ch4_lake_m2, col = "black",xlab = expression(atop("fitted lake (F"[CH4]*")"^(1/3)~"values",
  "(mg C m"^-2~"h"^-1*") from model 2")), ylab = "residuals" )
qqnorm(resid(ch4_lake_m2), main = "Q-Q plot")
qqline(resid(ch4_lake_m2), col = "red")

#variances
(0.04183/(0.04183+.13955+0+.34232))*100 #8% explained by chamber
(.13955/(0.04183+.13955+0+.34232))*100 # 27% explained by zone; 0% by lake
(.34232/(0.04183+.13955+0+.34232))*100 # 65% explained by fixed effects

#visualize table of outputs **data still transformed**
tab_model(ch4_lake_m2,
          show.p = FALSE, #no p-values 
  show.stat = TRUE, #t value  
  show.df = TRUE, #show degrees of freedom
 digits = 4)

```

likelihood ratios for CH4 Model 2 (no DO, pH)
```{r}
# full model
Form3 <- formula(ch4_cuberoot ~ tsurf_water_degc + cond_uscm.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no))
M2ch4.full <- lmer(Form3, data = lake_lmmdata, REML = F)

#test removal of fixed effects

#get rid of Twater NAs to test dfs of the same size
lake_noTNA <- lake_lmmdata[!is.na(lake_lmmdata$tsurf_water_degc),]
M2ch4.full.T <- lmer(ch4_cuberoot ~ tsurf_water_degc + cond_uscm.1+ disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no), data = lake_noTNA, REML = F)
M2ch4.A <- update(M2ch4.full.T, .~. -tsurf_water_degc) #remove Twater
anova(M2ch4.full.T, M2ch4.A)

#get rid of EC NAs to test dfs of the same size
lake_noECNA <- lake_lmmdata[!is.na(lake_lmmdata$cond_uscm.1),]
M2ch4.full.EC <- lmer(ch4_cuberoot ~ tsurf_water_degc + cond_uscm.1+ disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no), data = lake_noECNA, REML = F)
M2ch4.B <- update(M2ch4.full.EC, .~. -cond_uscm.1) #remove EC
  anova(M2ch4.full.EC, M2ch4.B)  

M2ch4.C <- update(M2ch4.full, .~. -disturbance) #remove disturbance
  anova(M2ch4.full, M2ch4.C)  

M2ch4.D <- update(M2ch4.full, .~. -area_km2) #remove area
  anova(M2ch4.full, M2ch4.D)

M2ch4.E <- update(M2ch4.full, .~. -depth_m) #remove depth
  anova(M2ch4.full, M2ch4.E)

```


Lake N2O LMM

```{r}
hist(lake_lmmdata$n2o_mg_m2_h)

#cube root transformation
lake_lmmdata$n2o_cuberoot <- sign(lake_lmmdata$n2o_mg_m2_h) * abs(lake_lmmdata$n2o_mg_m2_h)^(1/3)

print(lake_lmmdata$n2o_cuberoot)
hist(lake_lmmdata$n2o_cuberoot) 

#LMM with transformed n2o
lake_n2o_m1 <- lmer(n2o_cuberoot ~ tsurf_water_degc + cond_uscm.1 + ph + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no),
             data = lake_lmmdata, REML= T)
summary(lake_n2o_m1)

#examine residuals
plot(lake_n2o_m1, col = "black", xlab =expression("fitted lake (F"[N2O]*")"^(1/3)~"values (mg m"^-2~"h"^-1*")"), ylab = "residuals") #much better
qqnorm(resid(lake_n2o_m1), main = "Q-Q plot")
qqline(resid(lake_n2o_m1), col = "red")

#calculate variances
(0.009042/(0.009042+0.05528+0.08446))*100 #6% variance explained by zone; 0% by chamber
(0.05528/(0.009042+0.05528+0.08446))*100 #37% variance explained by site aka lake
(0.08446/(0.009042+0.05528+0.08446))*100 #57% explained by fixed effects

#visualize table of outputs **data still transformed**
tab_model(lake_n2o_m1,
          show.p = FALSE, #no p-values
  show.stat = TRUE, #t value
  show.df = TRUE, #show degrees of freedom
 digits = 4)

```

likelihood ratios for N2O LMM
```{r}
#full model
Form4 <- formula(n2o_cuberoot ~ tsurf_water_degc + cond_uscm.1 + ph + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no))
Mn2o.full <- lmer(Form4, data = lake_lmmdata, REML = F)

#test removal of fixed effects

Mn2o.C <- update(Mn2o.full, .~. -tsurf_water_degc) #remove Twater
  anova(Mn2o.full, Mn2o.C)
  
Mn2o.D <- update(Mn2o.full, .~. -cond_uscm.1) #remove conductivity
  anova(Mn2o.full, Mn2o.D)
 
#get rid of pH NAs to test dfs of the same size
lake_nophNA <- lake_lmmdata[!is.na(lake_lmmdata$ph),] #repeated from co2
Mn2o.full.ph <- lmer(n2o_cuberoot ~ tsurf_water_degc + cond_uscm.1 + ph + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no), data = lake_nophNA, REML = F)
Mn2o.E <- update(Mn2o.full.ph, .~. -ph) #remove ph
  anova(Mn2o.full.ph, Mn2o.E)
 
#get rid of DO NAs to test dfs of the same size
lake_nodoNA <- lake_lmmdata[!is.na(lake_lmmdata$do_mgl.1),] #repeated from co2
Mn2o.full.do <- lmer(n2o_cuberoot ~tsurf_water_degc + cond_uscm.1 + ph + do_mgl.1 + disturbance + area_km2 + depth_m +
             (1|site/sampling_zone/chamber_no), data = lake_nodoNA, REML = F)
Mn2o.F <- update(Mn2o.full.do, .~. -do_mgl.1) #remove DO
  anova(Mn2o.full.do, Mn2o.F)
 

Mn2o.G <- update(Mn2o.full, .~. -disturbance) #remove disturbance
  anova(Mn2o.full, Mn2o.G)
 
Mn2o.H <- update(Mn2o.full, .~. -area_km2) #remove area
  anova(Mn2o.full, Mn2o.H)

Mn2o.I <- update(Mn2o.full, .~. -depth_m) #remove depth
  anova(Mn2o.full, Mn2o.I)

```

